---
title: "RNAseq differential gene expression (DE) pilot test"
author: "John Kozlosky"
date: "2026-03-07"
output: html_document
---

# Setup

## R environment setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Required normal R packages
library(rstudioapi)
library(tidyr)
library(ggplot2)
library(dplyr)
library(readr)
library(tibble)
library(pheatmap)
library(ashr)
library(RColorBrewer)
library(matrixStats)
library(here)
library(plotly)
library(rlang)
library(AnnotationDbi)
library(enrichplot)

# Install bioconductor if missing
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.23")

# Required bioconductor packages
packages <- c("DESeq2", "biomaRt", "org.Hs.eg.db", "clusterProfiler", "ggupset", "msigdbr")

# Identify and install any missing bioconductor packages
# Load all required bioconductor packages
missing <- packages[!(packages %in% rownames(installed.packages()))]
if (length(missing)) BiocManager::install(missing)
lapply(packages, library, character.only = TRUE)

# Run data directory setup helper file
# Stores user's inputted data location in the data_dir variable
# Note: data_dir (really more like project directory) and data_folder (actually folder of data) are going to cause some confusion. Make this better at some point
source(file.path("setup_data_dir.R"))


# biomaRt setup
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")

```

## Data locations & assay config

```{r configuration_CHS, eval=TRUE}
# Hi there! Please specify the location/names of your raw count matrix
# and metadata files. Then specify the parameters for QC and DE analysis.

# Counts matrix should be RAW, UNADJUSTED COUNTS.
# DESeq2 adjusts for library size internally.

# TO-DO: Modify import code so I can specify sample ID and gene ID column names here

# data_folder is the subfolder of "data_dir" (project folder), which contains the 
# actual data for analysis
data_folder <- here(data_dir, "Data", "CHS")


# Data locations
file_countsMatrix <- here(data_folder, "CHS_G632_rawcounts.csv")
countsMatrx_skipLast <- 5 # Skip the last number of lines containing QC info

file_metadata <- here(data_folder, "CHS_G632_metadata.csv")
metadata_numSamples <- 40 # Temp - using this to keep only 40 samples in metadata file
# because I can't find the counts matrix for the rest!!

file_demographics <- here(data_folder, "chs_data_for_John.RData")

# QC setup
input_add1_allCounts = FALSE # Add 1 to all counts to "fix" the "Every gene contains at least one zero" error. This is a shitty way to do this
input_metadata_sort = "CM" # Column to sort metadata by. Set NULL to disable
input_QC_heatmap_var = "cm_cat_comp" # Hierarchical heatmap grouping
input_QC_PCA_intgroup = c("cm_cat_comp")
input_QC_PCA_multiPC_intgroup = c("CM", "sex") # Multi-PC PCA plot grouping
input_QC_PCA_3D_intgroup = expr(CM)
input_samples_drop = c(4,11, 12, 14, 40) #13 is unknown CM, rest are outliers(?)
input_samples_dropFilter = "CM = \"Unknown\""

# DE Setup
input_DE_design = ~ t1_age_g2 +	t1_gender_g2 +	t1_race_g2 +	t1_eth_g2 + t1_bmi + CM
#cm_cat_comp is categorical, CM is binary. Using CM for now bc I don't know how to set up multiple comparison groups in DESeq2 yet!! Idk how to ANOVA this shit
input_DE_contrast = c("CM", "Yes", "No")
input_DE_alpha = 0.05
input_DE_lfcThreshold = 0 # Datacamp used 0.32

# Graph setup
input_heatmap_colors = brewer.pal(n = 6, name = "YlOrRd")

# Gene attributes to pull from ensembl
input_gene_attributes <- c(
  "ensembl_gene_id",
  "ensembl_gene_id_version",
  "hgnc_symbol",
  "external_gene_name",
  "entrezgene_id",
  "gene_biotype",
  "description",
  "chromosome_name",
  "start_position",
  "end_position",
  "strand"
)
```

```{r configuration_stressStudy, eval = FALSE}
# Hi there! Please specify the location/names of your raw count matrix
# and metadata files. Then specify the parameters for QC and DE analysis.

# Counts matrix should be RAW, UNADJUSTED COUNTS.
# DESeq2 adjusts for library size internally.

# TO-DO: Modify import code so I can specify sample ID and gene ID column names here

# Data locations
# data_folder is the subfolder of "data_dir" (project folder), which contains the 
# actual data for analysis
data_folder <- here(data_dir, "Data", "Stress study")
file_countsMatrix <- here(data_folder,
                          "Counts_Combined_MM_20bp_Final.tsv")
countsMatrx_skipLast <- 0 # Skip the last number of lines containing QC info
file_metadata <- here(data_folder,
                      "stressStudy_subjectInfo_abridged.csv")
metadata_numSamples <- NULL # Only grab this many samples. Set NULL to disable.

# QC setup
input_add1_allCounts = TRUE # Add 1 to all counts to "fix" the "Every gene contains at least one zero" error. This is a shitty way to do this
input_metadata_sort = NULL # Column to sort metadata by. Set NULL to disable
input_QC_heatmap_var = c("session", "timepoint") # Hierarchical heatmap grouping
input_QC_PCA_intgroup = c("session") # PCA grouping
input_QC_PCA_multiPC_intgroup = c("session", "timepoint") # Multi-PC PCA plot grouping
input_QC_PCA_3D_intgroup = expr(session)
input_rawcounts_samples_drop = c(NULL) # Vector of sample numbers in rawcounts to drop. Set NULL to disable.
input_metadata_samples_filter = expr(group == 2) # Filter metadata samples to this. Set NULL to disable.

# DE Setup
input_DE_design = ~ participant + session * timepoint
input_DE_contrast = c("session", "control", "stress")
input_DE_alpha = 0.05
input_DE_lfcThreshold = 0 # Datacamp used 0.32

# Graph setup
input_heatmap_colors = brewer.pal(n = 6, name = "YlOrRd")

```

## Import data

```{r import_data_CHS, eval= TRUE}
# Import counts data
rawcounts <- read_csv(file_countsMatrix,
         col_types =  cols(
           .default = col_integer(),
           `...1` = col_character()
         )) %>%
  `colnames<-`(c("...1", 1:40)) %>%
  column_to_rownames(var = "...1") %>%
  dplyr::select(-all_of(input_samples_drop))

# Load participant demographics
load(file_demographics)

# Import sample metadata
metadata <- read_csv(file_metadata,
  col_types = cols(
    .default = col_guess(),
    "CM" = col_factor(),
    "Batch" = col_factor(),
    "g2id" = col_character()
  )) %>%
  head(-countsMatrx_skipLast) %>%
  head(metadata_numSamples) %>%
  column_to_rownames(var = "Sample_ID_1") %>%
  filter(!(rownames(.) %in% input_samples_drop))

if(!is.null(input_metadata_sort)) {metadata <- metadata %>% arrange(!!sym(input_metadata_sort))} # wtf does this do? sorting I think?


# Append demographics data to sample metadata
metadata <- left_join(metadata, chs_data_demo, by = "g2id")

```

```{r import_data_stress, eval = FALSE}
# Import raw counts data
rawcounts <- read_tsv(file_countsMatrix,
                      col_types = cols(
                       .default = col_integer(),
                       Geneid = col_character()
                     )) %>%
  replace(is.na(.), 0) %>%
  column_to_rownames(var = "Geneid") %>%
  select(-all_of(input_rawcounts_samples_drop))


# To-do:
# Instead of using the abridged datafile (manually created in excel),
# import and merge the two separate datafiles. Steps to do this
  # Import the ORIGINAL separate datafiles for participants 1 - 13 and 14 - 35
  # Transpose the datafile for participatns 1 - 13 (it's flipped the wrong way)
  # Duplicate IDs - 1x for control session, 1x for stress session
    # This may heave to be done differently for both datafiles
  # Quadruplicate IDs - 1x for each timepoint (4x total)
    # This effectively turns the PARTICIPANT level roster into a SAMPLE level roster
  # Merge both datafiles together by merging columns of the same name
  # Flag columns with no matching name to merge later, or
  # Just discard the un-matching columns for now
  # Ensure proper datatypes and calculate whatever else I need to

# This will yield us SAMPLE-level list that contains, per 1x participant
  # 8x total samples (split into 2x sessions)
    # 4x control session samples (split into 4x timepoints)
    # 4x stress session samples (split into 4x timepoints)



# import abridged metadata
# fix this by later by importing and merging both full participant datafiles
metadata <- read_csv(file_metadata,
                     col_types = cols(
                       .default = col_factor(),
                       participant = col_integer(),
                       `date...6` = col_date(format = "%m/%d/%Y"),
                       `date...10` = col_date(format = "%m/%d/%Y"),
                       birthday = col_date(format = "%m/%d/%Y"),
                       age = col_double()
                     )) %>%
  mutate(age = as.vector(scale(age))) %>%
  column_to_rownames(var = "sampleID") %>%
  filter(!!input_metadata_samples_filter) # Drop pre-specified samples
  


```

### Clean data

```{r import_data_clean_data}

# Drop samples that have NAs in columns specified by DESeq2 design formula
# 1. Extract variables from the design
design_cols <- all.vars(input_DE_design)
# 2. Filter metadata to keep only rows where NONE of these columns are NA
metadata <- metadata[complete.cases(metadata[, design_cols]), ]

# Z-score standardize numeric variables
metadata <- metadata %>%
  mutate(across(all_of(design_cols) & where(is.numeric), scale))

# Use the match() function to reorder the columns of the raw counts
# Find only rows from metadata that exist in cols of rawcounts
desired_order <- rownames(metadata)
rawcounts <- rawcounts[, desired_order[desired_order %in% colnames(rawcounts)], drop = FALSE]
# Drop samples from metadata that don't appear in the rawcounts so DESeq2 doesn't get upset
# And vice cersa - drop samples from rawcounts that don't appear in metadata (bc participant has been filtered out due to NAs)
metadata <- metadata[rownames(metadata) %in% colnames(rawcounts), , drop = FALSE]

```

## DESeq2 dataset prep

```{r DE_dataset_prep}

# QC - Remove genes with no reads!!

# Flag genes with 0 reads, low reads (1 - 10), then filter out from rawcounts
genes_ZeroReads <- rownames(rawcounts[rowSums(rawcounts) == 0, ])
genes_LowReads <- rownames(rawcounts[rowSums(rawcounts) < 10, ])
genes_ToRemove <- c(genes_ZeroReads, genes_LowReads)

rawcountsFiltered <- rawcounts[!(row.names(rawcounts) %in% genes_ToRemove), ]


# If specified -
# Add 1 read to every gene/sample
# Only way I can get stress study data to work :(
if (input_add1_allCounts == TRUE) {rawcounts <- rawcounts + 1}


dds <- DESeqDataSetFromMatrix(
  countData = rawcountsFiltered,
  colData = metadata,
  design = input_DE_design
)
```

# QC

## Hierarchical heatmap

```{r QC_hierarchical_heatmap}
# Estimate size factors
dds <- estimateSizeFactors(dds)
# Normalize counts to library size (for heatmap)
dds_normalized_counts <- counts(dds, normalized = TRUE)
# Transform counts with variance stabilized dispersion (what does this do?)
vsd_dds <-vst(dds, blind = TRUE)
# Pull out matrix of vsd counts
vsd_matrix_dds <- assay(vsd_dds)
# Compute correlations between samples
vsd_cor_dds <- cor(vsd_matrix_dds)
pheatmap(vsd_cor_dds, annotation = dplyr::select(metadata, input_QC_heatmap_var),
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = "QC: Sample expression correlation heatmap")
```

Samples from the same group should cluster together. Different groups of samples should cluster apart.

## PCA

### PC1 vs. PC2 plot

```{r QC_PCA_plot}
plotPCA(vsd_dds, intgroup = input_QC_PCA_intgroup, ntop = 500, returnData = FALSE)
```

Your specified condition of interest would ideally be PC1 or PC2. You should expect to see separation of your groups of interest of other known sources or significant variability.

### Scree plot

```{r QC_PCA_scree_plot}

# From ChatGPT - review code later
ntop <- 500 # DESeq defaults to 500
rv <- matrixStats::rowVars(assay(vsd_dds))
select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]

pca <- prcomp(t(assay(vsd_dds)[select, ]))   # same call DESeq2 uses
percentVar <- (pca$sdev^2) / sum(pca$sdev^2)

scree <- data.frame(
  PC = seq_along(percentVar),
  percent = 100 * percentVar
)

ggplot(scree, aes(x = PC, y = percent, fill = percent)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c(option = "A") +
  # coord_cartesian(ylim = c(0,100)) +
  labs(x = "Principal component", y = "% variance explained")

```

### PC1 - PC4 multi-plot

```{r PCA_3D}
# scores for first 4 PCs
pc_scores <- as.data.frame(pca$x[, 1:3, drop = FALSE])
pc_scores$sample <- rownames(pc_scores)

# attach metadata you want to color/shape by (edit to match your colData columns)
pca_meta <- as.data.frame(colData(vsd_dds))

pc_scores <- cbind(pc_scores, pca_meta)

labels <- paste0("PC", 1:3, " (", round(percentVar[1:3]*100), "%)")

plot_ly(pc_scores, 
        x = ~PC1, y = ~PC2, z = ~PC3, 
        color = new_formula(NULL, input_QC_PCA_3D_intgroup), # Make this changable with config
        type = 'scatter3d', 
        mode = 'markers') %>%
  layout(scene = list(
    xaxis = list(title = labels[1]),
    yaxis = list(title = labels[2]),
    zaxis = list(title = labels[3])
  ))


```

# DE modeling

## Processing

```{r DE_processing}
dds <- DESeq(dds)
```

## Dispersion calculation

```{r QC_dispersions}
plotDispEsts(dds)
```

Samples should generally follow and cluster around the fitted line. Dispersions should generally decease with increasing mean.

# Results

```{r results}
dds_results <- results(dds,
                       contrast = input_DE_contrast,
                       alpha = input_DE_alpha,
                       lfcThreshold = input_DE_lfcThreshold)

dds_results <- lfcShrink(dds,
                         contrast = input_DE_contrast,
                         res = dds_results,
                         type = "ashr")
# Datacamp uses "normal" lfc shrink algorithm from Love et al, 2014.
# "ashr" from Stephens, 2016 yields less biased results

summary(dds_results)

# Save all results as dataframe, flag significant results,
# then filter to significant results
dds_results_all <- data.frame(dds_results) %>%
  mutate(threshold = padj < input_DE_alpha)

dds_results_sig <- subset(dds_results_all, threshold == TRUE) %>% arrange(padj)

```

# QC (post-model)

## MA plot

```{r QC_MA_plot}
# Scatterplot of normalized counts (x) vs. log2 fold changes (y)
plotMA(dds_results)
```

## Volcano plot

```{r QC_volcano_plot}
ggplot(dds_results_all) + 
        geom_point(aes(x = log2FoldChange, y = -log10(padj), color = threshold)) + 
        xlab("log2 fold change") + 
        ylab("-log10 adjusted p-value") + 
        theme(legend.position = "right", 
              plot.title = element_text(size = rel(1.5), hjust = 0.5), 
              axis.title = element_text(size = rel(1.25)))
```

## Significant genes heatmap

```{r QC_heatmp_significant_genes}
dds_normalized_counts_sig <- dds_normalized_counts[rownames(dds_results_sig), ]

pheatmap(dds_normalized_counts[select, ], # No significantly DE genes, so choosing top 500 variation for now. change back later
         color = input_heatmap_colors, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         show_rownames = FALSE,
         annotation =dplyr:: select(metadata, input_QC_PCA_intgroup), 
         scale = "row",
         main = "QC: Gene-level expression level heatmap")

```

# Gene annotation

```{r ensembl_annotation, eval = FALSE}
  
# Note: Ensembl is kind of fucked rtight now
# because they're updating the website.
# It's been like this for 3 months, apparently. So I'm not going to
# fuck with ensembl. I'm converting the ensembl IDs to entrez IDs
# annd going from there.

# Should we keep ensembl IDs as the rownames here?
# Or just let them be removed (converted to column) with
# rownames_to_column()?
dds_results_all_annotated <- dds_results_all %>% 
  mutate(
    ensembl_gene_id = sub("\\..*$", "", rownames(dds_results_all))
  )

ensembl <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl",
  mirror = "useast"
)


geneInfo <- getBM(
  attributes = input_gene_attributes,
  filters = "ensembl_gene_id",
  values = unique(res_tbl$ensembl_gene_id),
  mart = ensembl
) %>%
  mutate(entrezgene_id = as.character(entrezgene_id)) %>%
  distinct(ensembl_gene_id, .keep_all = TRUE)

dds_results_all_annotated <- dds_results_all_annotated %>%
  left_join(geneInfo, by = "ensembl_gene_id")
```

```{r annotation}

# Using the ensembl ID, map the following info per gene:
# gene symbol, entrez ID, gene name

# Convert results into regular table
dds_results_all_table <- as.data.frame(dds_results_all) %>%
  rownames_to_column("ensembl_gene_id_raw") %>%
  mutate(
    ensembl_gene_id = sub("\\..*$", "", ensembl_gene_id_raw)
  )

# Locally match Ensembl IDs to symbols, Entrez IDs, and names
genes_annotated <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(res_tbl$ensembl_gene_id),
  keytype = "ENSEMBL",
  columns = c("ENSEMBL", "SYMBOL", "ENTREZID", "GENENAME")
) %>%
  as_tibble() %>%
  distinct()

# Merge annotation info into results table
dds_results_all_table_annotated <- dds_results_all_table %>%
  left_join(genes_annotated, by = c("ensembl_gene_id" = "ENSEMBL"), keep = TRUE)
```

# Over-representation analysis

```{r analysis_ORA}

# Significantly DE genes
sig_genes_ensembl <- dds_results_all_table_annotated %>%
  filter(!is.na(padj), padj < 0.05, !is.na(ENSEMBL)) %>%
  pull(ENSEMBL) %>%
  unique()

# All genes
background_genes_ensembl <- dds_results_all_table_annotated %>%
  filter(!is.na(ENSEMBL)) %>%
  pull(ENSEMBL) %>%
  unique()

# ORA for all 3 ontology options:
# biological processes, molecular functions, and cellular components
ego_ora_bp <- enrichGO(
  gene = sig_genes_ensembl,
  universe = background_genes_ensembl,
  OrgDb = org.Hs.eg.db,
  keyType = "ENSEMBL",
  ont = "ALL"
)

# Separate analyses for individual components
# Leaving these in here but commented out for now
# It might be helpful later to run these separately when we 
# have more data

# ORA for biological processes
# ego_ora_bp <- enrichGO(
#   gene = sig_genes_ensembl,
#   universe = background_genes_ensembl,
#   OrgDb = org.Hs.eg.db,
#   keyType = "ENSEMBL",
#   ont = "BP"
# )
# 
# # ORA for molecular functions
# ego_ora_mf <- enrichGO(
#   gene = sig_genes_ensembl,
#   universe = background_genes_ensembl,
#   OrgDb = org.Hs.eg.db,
#   keyType = "ENSEMBL",
#   ont = "MF"
# )
# 
# # ORA for cellular components
# ego_ora_cc <- enrichGO(
#   gene = sig_genes_ensembl,
#   universe = background_genes_ensembl,
#   OrgDb = org.Hs.eg.db,
#   keyType = "ENSEMBL",
#   ont = "CC"
# )

# Helper function used below
ratio_to_num <- function(x) {
  sapply(strsplit(x, "/"), function(z) as.numeric(z[1]) / as.numeric(z[2]))
}

# Convert result tables to dataframes
# Note - write another function to do this 3x 
# or just do ont = "ALL" and add a column indicating BP, MF, or CC
# since there aren't many results rn

# Get usable results dataframe from results object
# Currently baked into a function so I can re-use it later
# if I'm going to run the ontologies separately
getDF_ego_ora <- function(ego_ora_results) {
  ego_ora_df <- as.data.frame(ego_ora_results)

  # Convert expression ratios to numbers - if no data, skip
  if(nrow(ego_ora_results) > 0) {
    ego_ora_df <- ego_ora_df %>%
      mutate(
        GeneRatio_num = ratio_to_num(GeneRatio),
        BgRatio_num = ratio_to_num(BgRatio),
        FoldEnrichment = GeneRatio_num / BgRatio_num
      ) %>%
      arrange(p.adjust)
  }
    
  dfVarName = paste0(deparse(substitute(ego_ora_results)), "_df")
  
  assign(dfVarName,
          ego_ora_df,
         envir = .GlobalEnv)
  
  return(get(dfVarName))
}

# Extract results
getDF_ego_ora(ego_ora_all)


#ChatGPT QA checks
length(sig_genes_ensembl)
length(background_genes_ensembl)

sum(duplicated(sig_genes_ensembl))
sum(duplicated(background_genes_ensembl))

length(intersect(sig_genes_ensembl, background_genes_ensembl))
setdiff(sig_genes_ensembl, background_genes_ensembl) |> head()

mapped_check <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = background_genes_ensembl,
  keytype = "ENSEMBL",
  columns = c("ENSEMBL", "ENTREZID", "SYMBOL")
) %>%
  as_tibble()

mapping_rate <- mapped_check %>%
  summarise(
    input_genes = length(unique(background_genes_ensembl)),
    mapped_ensembl = n_distinct(ENSEMBL[!is.na(ENTREZID)]),
    mapping_fraction = mapped_ensembl / input_genes
  )

mapping_rate

```

## ORA results visualization

### Dot plot

```{r ora_vis_dotplot}
# To-do: make the split by ontology clearer,
# or just facet by ontology
ego_ora_all_reordered <- ego_ora_all %>% arrange(desc(ONTOLOGY), desc(Count))
dotplot(ego_ora_all_reordered, showCategory = 20, split = "ONTOLOGY")
```

### Bar plot

```{r ora_vis_barplot}

p_nofacet <- barplot(
  ego_ora_all_reordered,
  x = "Count",
  showCategory = 15,
  split = "ONTOLOGY"
) +
  aes(fill = ONTOLOGY) +
  scale_fill_brewer(palette = "Set2")

p_nofacet +
  geom_text(aes(label = Count), nudge_x = 2) +
  scale_y_discrete()

```

# Gene set enrichment analysis

```{r analysis_GSEA}

```
