---
title: "RNAseq differential gene expression (DE) pilot test"
author: "John Kozlosky"
date: "2026-03-07"
output: html_document
---

# Setup

## R environment setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Required normal R packages
library(tidyr)
library(ggplot2)
library(dplyr)
library(readr)
library(tibble)
library(pheatmap)
library(ashr)
library(RColorBrewer)
library(matrixStats)
library(here)
library(plotly)
library(rlang)

# Install bioconductor if missing
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Required bioconductor packages
packages <- c("DESeq2")

# Identify missing packages,
# install all missing packages at once
# Load all missing packages
missing <- packages[!(packages %in% rownames(installed.packages()))]
if (length(missing)) BiocManager::install(missing)
lapply(packages, library, character.only = TRUE)

```

## Data locations & assay config

```{r configuration_CHS, eval=FALSE, class.source="fold-hide", include=FALSE}
# Hi there! Please specify the location/names of your raw count matrix
# and metadata files. Then specify the parameters for QC and DE analysis.

# Counts matrix should be RAW, UNADJUSTED COUNTS.
# DESeq2 adjusts for library size internally.

# TO-DO: Modify import code so I can specify sample ID and gene ID column names here

data_folder <- here("Data", "CHS")


# Data locations
file_countsMatrix <- "CHS_G632_rawcounts.csv"
countsMatrx_skipLast <- 5 # Skip the last number of lines containing QC info
file_metadata <- "CHS_G632_metadata.csv"
metadata_numSamples <- 40 # Temp - using this to keep only 40 samples in metadata file
# because I can't find the counts matrix for the rest!!

# QC setup
input_metadata_sort = "CM" # Column to sort metadata by. Set NULL to disable
input_QC_heatmap_var = "CM" # Hierarchical heatmap grouping
input_QC_PCA_intgroup = c("CM")
input_samples_drop = c(4,11, 12, 14, 40) #13 is unknown CM, rest are outliers(?)
input_samples_dropFilter = "CM = \"Unknown\""

# DE Setup
input_DE_design = ~ CM # Having trouble running Batch + CM. Running just CM for now.
input_DE_contrast = c("CM", "Yes", "No")
input_DE_alpha = 0.05
input_DE_lfcThreshold = 0 # Datacamp used 0.32

# Graph setup
input_heatmap_colors = brewer.pal(n = 6, name = "YlOrRd")
```

```{r configuration_stressStudy, echo = FALSE}
# Hi there! Please specify the location/names of your raw count matrix
# and metadata files. Then specify the parameters for QC and DE analysis.

# Counts matrix should be RAW, UNADJUSTED COUNTS.
# DESeq2 adjusts for library size internally.

# TO-DO: Modify import code so I can specify sample ID and gene ID column names here

# Data locations
data_folder <- here("Data", "Stress study")
file_countsMatrix <- here(data_folder,
                          "Counts_Combined_MM_20bp_Final.tsv")
countsMatrx_skipLast <- 0 # Skip the last number of lines containing QC info
file_metadata <- here(data_folder,
                      "stressStudy_subjectInfo_abridged.csv")
metadata_numSamples <- NULL # Only grab this many samples. Set NULL to disable.

# QC setup
input_metadata_sort = NULL # Column to sort metadata by. Set NULL to disable
input_QC_heatmap_var = c("session", "timepoint") # Hierarchical heatmap grouping
input_QC_PCA_intgroup = c("session", "timepoint") # PCA grouping
input_QC_PCA_multiPC_intgroup = c("session", "timepoint") # Multi-PC PCA plot grouping
input_QC_PCA_3D_intgroup = expr(session)
input_rawcounts_samples_drop = c(NULL) # Vector of sample numbers in rawcounts to drop. Set NULL to disable.
input_metadata_samples_filter = expr(group == 2) # Filter metadata samples to this. Set NULL to disable.

# DE Setup
input_DE_design = ~ sex + age + session + timepoint
input_DE_contrast = c("session", "control", "stress")
input_DE_alpha = 0.05
input_DE_lfcThreshold = 0 # Datacamp used 0.32

# Graph setup
input_heatmap_colors = brewer.pal(n = 6, name = "YlOrRd")

```

## Import data

```{r import_data_CHS, eval=FALSE, include=FALSE}
# Import counts data
rawcounts <- read_csv(file_countsMatrix,
         col_types =  cols(
           .default = col_integer(),
           `...1` = col_character()
         )) %>%
  `colnames<-`(c("...1", 1:40)) %>%
  column_to_rownames(var = "...1") %>%
  select(-all_of(input_samples_drop))

# Import sample metadata
metadata <- read_csv(file_metadata,
  col_types = cols(
    .default = col_guess(),
    "CM" = col_factor(),
    "Batch" = col_factor()
  )) %>%
  head(-countsMatrx_skipLast) %>%
  head(metadata_numSamples) %>%
  column_to_rownames(var = "Sample_ID_1") %>%
  filter(!(rownames(.) %in% input_samples_drop))

if(!is.null(input_metadata_sort)) {metadata <- metadata %>% arrange(!!sym(input_metadata_sort))}

# Use the match() function to reorder the columns of the raw counts
cols_to_sort <- rownames(metadata) %in% colnames(rawcounts)
reorder_idx <- match(rownames(metadata), colnames(rawcounts))
rawcounts <- rawcounts[ , reorder_idx]

```

```{r import_data_stress, echo = TRUE}
# Import raw counts data
rawcounts <- read_tsv(file_countsMatrix,
                      col_types = cols(
                       .default = col_integer(),
                       Geneid = col_character()
                     )) %>%
  replace(is.na(.), 0) %>%
  column_to_rownames(var = "Geneid") %>%
  select(-all_of(input_rawcounts_samples_drop))

# import sample metadata
metadata <- read_csv(file_metadata,
                     col_types = cols(
                       .default = col_factor(),
                       participant = col_integer(),
                       `date...6` = col_date(format = "%m/%d/%Y"),
                       `date...10` = col_date(format = "%m/%d/%Y"),
                       birthday = col_date(format = "%m/%d/%Y"),
                       age = col_double()
                     )) %>%
  mutate(age = as.vector(scale(age))) %>%
  column_to_rownames(var = "sampleID") %>%
  filter(!!input_metadata_samples_filter)
  

# Use the match() function to reorder the columns of the raw counts
# Find only rows from metadata that mexist in cols of rawcounts
desired_order <- rownames(metadata)
rawcounts <- rawcounts[, desired_order[desired_order %in% colnames(rawcounts)], drop = FALSE]
# Drop samples from metadata that don't appear in the rawcounts so DESeq2 doesn't get upset
metadata <- metadata[rownames(metadata) %in% colnames(rawcounts), , drop = FALSE]
```

## DESeq2 dataset prep

```{r DE_dataset_prep}

# QC - Remove genes with no reads
rawcounts <- rawcounts[rowSums(rawcounts) >= 0, ]
# Add 1 read to every gene/sample
# Only way I can get stress study data to work :(
rawcounts <- rawcounts + 1

dds <- DESeqDataSetFromMatrix(
  countData = rawcounts,
  colData = metadata,
  design = input_DE_design
)
```

# QC

## Hierarchical heatmap

```{r QC_hierarchical_heatmap}
# Estimate size factors
dds <- estimateSizeFactors(dds)
# Normalize counts to library size (for heatmap)
dds_normalized_counts <- counts(dds, normalized = TRUE)
# Transform counts with variance stabilized dispersion (what does this do?)
vsd_dds <-vst(dds, blind = TRUE)
# Pull out matrix of vsd counts
vsd_matrix_dds <- assay(vsd_dds)
# Compute correlations between samples
vsd_cor_dds <- cor(vsd_matrix_dds)
pheatmap(vsd_cor_dds, annotation = select(metadata, input_QC_heatmap_var),
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = "QC: Sample expression correlation heatmap")
```

Samples from the same group should cluster together. Different groups of samples should cluster apart.

## PCA

### PC1 vs. PC2 plot

```{r QC_PCA_plot}
plotPCA(vsd_dds, intgroup = input_QC_PCA_intgroup, ntop = 500, returnData = FALSE)
```

Your specified condition of interest would ideally be PC1 or PC2. You should expect to see separation of your groups of interest of other known sources or significant variability.

### Scree plot

```{r QC_PCA_scree_plot}

# From ChatGPT - review code later
ntop <- 500 # DESeq defaults to 500
rv <- matrixStats::rowVars(assay(vsd_dds))
select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]

pca <- prcomp(t(assay(vsd_dds)[select, ]))   # same call DESeq2 uses
percentVar <- (pca$sdev^2) / sum(pca$sdev^2)

scree <- data.frame(
  PC = seq_along(percentVar),
  percent = 100 * percentVar
)

ggplot(scree, aes(x = PC, y = percent, fill = percent)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c(option = "A") +
  # coord_cartesian(ylim = c(0,100)) +
  labs(x = "Principal component", y = "% variance explained")

```

### PC1 - PC4 multi-plot

```{r PCA_3D}
# scores for first 4 PCs
pc_scores <- as.data.frame(pca$x[, 1:3, drop = FALSE])
pc_scores$sample <- rownames(pc_scores)

# attach metadata you want to color/shape by (edit to match your colData columns)
meta <- as.data.frame(colData(vsd_dds))

pc_scores <- cbind(pc_scores, meta)

labels <- paste0("PC", 1:3, " (", round(percentVar[1:3]*100), "%)")

plot_ly(pc_scores, 
        x = ~PC1, y = ~PC2, z = ~PC3, 
        color = new_formula(NULL, input_QC_PCA_3D_intgroup), # Make this changable with config
        type = 'scatter3d', 
        mode = 'markers') %>%
  layout(scene = list(
    xaxis = list(title = labels[1]),
    yaxis = list(title = labels[2]),
    zaxis = list(title = labels[3])
  ))


```

# DE modeling

## Processing

```{r DE_processing}
dds <- DESeq(dds)
```

## Dispersion calculation

```{r QC_dispersions}
plotDispEsts(dds)
```

Samples should generally follow and cluster around the fitted line. Dispersions should generally decease with increaseing mean.

# Results

```{r results}
dds_results <- results(dds,
                       contrast = input_DE_contrast,
                       alpha = input_DE_alpha,
                       lfcThreshold = input_DE_lfcThreshold)

dds_results <- lfcShrink(dds,
                         contrast = input_DE_contrast,
                         res = dds_results,
                         type = "ashr")
# Datacamp uses "normal" lfc shrink algorithm from Love et al, 2014.
# "ashr" from Stephens, 2016 yields less biased results

summary(dds_results)

# Save all results as dataframe, flag significant results,
# then filter to significant results
dds_results_all <- data.frame(dds_results) %>%
  mutate(threshold = padj < input_DE_alpha)

dds_results_sig <- subset(dds_results_all, threshold == TRUE)

```

# QC (post-model)

## MA plot

```{r QC_MA_plot}
# Scatterplot of normalized counts (x) vs. log2 fold changes (y)
plotMA(dds_results)
```

## Volcano plot

```{r QC_volcano_plot}
ggplot(dds_results_all) + 
        geom_point(aes(x = log2FoldChange, y = -log10(padj), color = threshold)) + 
        xlab("log2 fold change") + 
        ylab("-log10 adjusted p-value") + 
        theme(legend.position = "right", 
              plot.title = element_text(size = rel(1.5), hjust = 0.5), 
              axis.title = element_text(size = rel(1.25)))
```

## Significant genes heatmap

```{r QC_heatmp_significant_genes}
dds_normalized_counts_sig <- dds_normalized_counts[rownames(dds_results_sig), ]

pheatmap(dds_normalized_counts[select, ], # No significantly DE genes, so choosing top 500 variation for now. change back later
         color = input_heatmap_colors, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         show_rownames = FALSE,
         annotation = select(metadata, session), 
         scale = "row",
         main = "QC: Gene-level expression level heatmap")

```
